### 2022-04-04 paper/localpsf.tex
working through next round of paper comments by Noemi

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(v) = \left(\Aop v\right)(u)$ for all $u,v \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure. After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representations of $\Aop$ and $\Aop^T$, but cannot easily compute individual matrix entries.

@@

Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
\begin{equation*}
y = f(m, u(m)) + \text{noise},
\end{equation*}
which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
\begin{equation}
\label{eq:state_pde}
0 = g(m,u).
\end{equation}
Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
\begin{equation}
\label{eq:minimization_problem}
\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
\end{equation}
where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$ \cite{LAPLACEAPPROXIMATION}.

@@

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. In this section we describe how to extend the action of $\Aop$ to distributions, which allows us to define the impulse responses (Section \ref{sec:impulse_response_2}). Then we state the conditions on $\Aop$ that our method requires (Section \ref{sec:conditions_2}). Finally, we detail finite element discretization (Section \ref{sec:finite_element_kernel}).
%We write $L^2(\Omega)'$ to denote the topological dual of $L^2(\Omega)$. 
%The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

@@

The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy,
	\end{align}
\end{subequations}
%\begin{align*}
%\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
%&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
%&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
%\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the last line in the derivation is well-defined. Hence, we use the last line of this derivation to \emph{define} the action of $\Aop$ on the distribution $\rho$ as follows:
\begin{equation}
\label{eq:action_on_distribution}
\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
\end{equation}
Since $\genericdistribution^*$ may not exist, here ``$\Aop \genericdistribution^*$'' is simply a notation that means ``the action of $\Aop$ on the distribution $\rho$,'' per \eqref{eq:action_on_distribution}.

>>>>

By ``matrix free,'' we mean that we have a black-box computational procedure through which we may apply $\Aop$ and its transpose\footnote{Recall that the transpose $\Aop^T:L^2(\Omega)\rightarrow L^2(\Omega)'$ is the unique operator satisfying $\left(\Aop u\right)(v) = \left(\Aop v\right)(u)$ for all $u,v \in L^2(\Omega)$, where the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$, and similar for the analogous operations with $\Aop^T$. }, $\Aop^T$, to arbitrary functions,
\begin{equation}
\label{eq:eval_maps}
u \mapsto\Aop u \qquad \text{and} \qquad v \mapsto \Aop^Tv,
\end{equation} 
but we do not have any further access to $\Aop$. In particular, we cannot easily access entries of $\Aop$'s integral kernel. In applications, evaluating the maps in \eqref{eq:eval_maps} may involve iteratively solving a large linear system, timestepping, or performing some other costly computational procedure. 
%After discretization, $\Aop$ becomes a dense matrix that is typically too large to be built and stored. This property is called ``matrix-free'' because at the discrete level it means one can compute matrix-vector products with the matrix representations of $\Aop$ and $\Aop^T$, but cannot easily compute individual matrix entries.

@@

Our motivation is Hessian approximation in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field from noisy observations of a state variable that depends on the parameter implicitly through the solution of a PDE.
In the deterministic approach to inverse problems, one solves a regularized minimization problem to find the parameter that best fits the data. The objective function in the minimization problem is defined implicitly, so information about the Hessian of the objective function is only accessible via application of the Hessian to vectors \cite{HESSIANACTION}. 
%In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. 
More accessible Hessian approximations are highly desirable because they permit fast solution of the minimization problem via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for the parameter \cite{LAPLACEAPPROXIMATION}.

%Our motivation for this work is approximation of Hessians in distributed parameter inverse problems governed by partial differential equations (PDEs). That is, inverse problems in which one seeks to reconstruct an unknown parameter field, $m$, from noisy observations, 
%\begin{equation*}
%y = f(m, u(m)) + \text{noise},
%\end{equation*}
%which depend on  a state variable $u$. In turn, $u$ depends on $m$ implicitly through the solution of a PDE, which we write generically as follows:
%\begin{equation}
%\label{eq:state_pde}
%0 = g(m,u).
%\end{equation}
%Here $u(m)$ denotes the solution of the PDE \eqref{eq:state_pde} as a function of $m$. In the deterministic approach to inverse problems, one typically finds $m$ as the solution to the minimization problem 
%\begin{equation}
%\label{eq:minimization_problem}
%\min_m \quad \frac{1}{2}\|y - f(m,u(m))\|_W^2 + \mathcal{R}(m),
%\end{equation}
%where $\|\cdot\|_W$ is weighted norm which depends on the noise covariance, and $\mathcal{R}(m)$ is a regularization term. The objective function in optimization problem \eqref{eq:minimization_problem} is defined indirectly via the implicit function theorem. As a result, information about the Hessian of the objective function is only accessible via the action of the Hessian on vectors. In particular, applying the Hessian to a vector requires solving two linear PDEs which take the form of linearizations of \eqref{eq:state_pde} \cite{HESSIANACTION}. More accessible approximations of the Hessian are highly desirable, because they allow for fast solution of \eqref{eq:minimization_problem} via Newton-type methods. More accessible Hessian approximations are also central to many methods for uncertainty quantification in Bayesian statistical approaches to the inverse problem, because the inverse of the Hessian locally approximates the Bayesian posterior covariance for $m$ \cite{LAPLACEAPPROXIMATION}.

@@

%Product-convolution (PC) approximations can (where applicable) approximate high rank operators efficiently.

@@

Let $\Omega \subset \mathbb{R}^d$ be a bounded domain (in applications typically $\gdim=1$, $2$, or $3$). We seek to approximate integral operators $\Aop:L^2(\Omega)\rightarrow L^2(\Omega)'$ that may be written in the following form:
\begin{equation}
\label{eq:kernel_representation}
(\Aop u)(v) := \int_\Omega \int_\Omega v(y) \Aker(y,x) u(x) dx dy.
\end{equation}
Here the linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, the scalar $\left(\Aop u\right)(v)$ is the result of applying that linear functional to $v \in L^2(\Omega)$.
The function $\Aker:\Omega \times \Omega \rightarrow \mathbb{R}$ is the integral kernel, which exists in principle but is not easily accessible to us. In this section we describe how to extend the action of $\Aop$ to distributions, which allows us to define the impulse responses (Section \ref{sec:impulse_response_2}). Then we state the conditions on $\Aop$ that our method requires (Section \ref{sec:conditions_2}). Finally, we detail finite element discretization (Section \ref{sec:finite_element_kernel}).
%We write $L^2(\Omega)'$ to denote the topological dual of $L^2(\Omega)$. 
%The linear functional $\Aop u \in L^2(\Omega)'$ is the result of applying $\Aop$ to $u\in L^2(\Omega)$, and $\left(\Aop u\right)(v)$ is the scalar that results from applying that linear functional to $v \in L^2(\Omega)$.

@@

The action of $\Aop$ may be extended to distributions if the kernel $\Phi$ is sufficiently regular. We derive the extension as follows. Suppose $\rho \in L^2(\Omega)'$, and let $\rho^* \in L^2(\Omega)$ denote the Riesz representative of $\rho$ with respect to the $L^2(\Omega)$ inner product. We have
\begin{subequations}
	\label{eq:extension_to_distributions}
	\begin{align}
		\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx ~dy \\
		&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy, \label{eq:action_on_distribution}
	\end{align}
\end{subequations}
%\begin{align*}
%\left(\Aop \genericdistribution^*\right)(v) &= \int_\Omega \int_\Omega v(y) \Phi(y,x) \genericdistribution^*(x) dx dy \\
%&= \int_\Omega v(y) \int_\Omega \Phi(y,x) \genericdistribution^*(x) dx \\
%&= \int_\Omega v(y) \genericdistribution\left(\Phi(y,\cdot)\right) dy 
%%= \left(v,\genericdistribution\left(\Phi(y,\cdot)\right)\right)_{L^2(\Omega)},
%\end{align*}
where $\Phi(y,\cdot)$ denotes the function $x \mapsto \Phi(y,x)$.
Now let $\mathcal{D}(\Omega) \subset L^2(\Omega)$ be a suitable space of test functions, and instead suppose $\rho:\mathcal{D}(\Omega) \rightarrow \mathbb{R}$ is a distribution. In this case, the Riesz representative $\rho^*$ may not exist, so the derivation in \eqref{eq:extension_to_distributions} is not valid. However, if $\Phi$ is sufficiently regular such that the function $y \mapsto \rho\left(\Phi(y,~\cdot~)\right)$ is well-defined for almost all $y\in\Omega$, and if this function is in $L^2(\Omega)$, then the the right hand side of \eqref{eq:action_on_distribution} is well-defined. Hence, we \emph{define} the action of $\Aop$ on the distribution $\rho$ to be the right hand side of \eqref{eq:action_on_distribution}. For convenience of notation, we denote this action by ``$\Aop \genericdistribution^*$,'' even if $\rho^*$ does not exist.
%\begin{equation}
%\label{eq:action_on_distribution}
%\left(\Aop \genericdistribution^*\right)(v) := \int_\Omega v(y) \genericdistribution\left(\Phi(y, \cdot)\right) dy.
%\end{equation}
%Since $\genericdistribution^*$ may not exist if $\rho$ is a distribution, 
%We write ``$\Aop \genericdistribution^*$'' to denote ``the action of $\Aop$ on the distribution $\rho$,'' as defined by \eqref{eq:action_on_distribution}.
